% !TeX spellcheck = en_GB

\documentclass[proc]{edpsmath}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{mathtools}

\newcommand{\bR}{\mathbb{R}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\vp}{v_\parallel}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\Dt}{\Delta t \;}

\newcommand{\dbx}{\dot{\mathbf{x}}}
\newcommand{\diff}[2]{\frac{\text{d} #1}{\text{d} #2}}
\newcommand{\pa}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ap}{a_\parallel}
\newcommand{\mean}[1]{\langle #1 \rangle}
\newcommand{\Bap}{B_\parallel^\ast}
\newcommand{\bracket}[2]{\left\{ #1, #2 \right\}}
\newcommand{\gcbracket}[2]{\left\{ #1, #2 \right\}_{\text{g.c.}}}
\newcommand{\dt}{\frac{\text{d}}{\text{d} t}}

\renewcommand{\d}{\;\text{d}}

%TODO: agreed?
\setlength{\parindent}{0pt}

\begin{document}

\title{SILAS - Semi-Lagrangian scheme with Arakawa splitting}
\thanks{CEMRACS}\thanks{Max-Planck}% At most 5 thanks

\author{Dominik Bell}\address{Max-Planck-Institut für Plasmaphysik, Garching, Germany; \email{dominik.bell@ipp.mpg.de\ \&\ frederik.schnack@ipp.mpg.de\ \&\ martin.campos-pinto@ipp.mpg.de\ \&\ sonnen@ipp.mpg.de}}\secondaddress{Technische Universität München, Zentrum Mathematik, Garching, Germany}
\author{Martin Campos Pinto}\sameaddress{1}
\author{Davor Kumozec}\address{Faculty of Sciences, University of Novi Sad, Serbia; \email{davor.kumozec@dmi.uns.ac.rs}}
\author{Frederik Schnack}\sameaddress{1}
\author{Eric Sonnendrücker}\sameaddress{1,2}


\begin{abstract}
In practice, there are various ways to solve gyro-kinetic equations. Here, we will combine the Semi-Lagrangian (SL) method and the Arakawa (AKW) scheme combined with a time-integrator. Both methods are successfully used in practice when applied individually, in our case, we combine them by first decomposing the problem into a fast (parallel) and a slow (perpendicular) dynamical system. The SL approach will be used to solve the fast and the AKW scheme for the slow subsystem. So far in our reference code \cite{pygyro_code}, the entire model is solved only using the SL method. Our goal is to replace the discretization of the slow subsystem by the AKW scheme, in order to improve preservation of conserved quantities of the physical system at hand. \\
After an introduction of the gyro-kinetic equations in Section \ref{sec:introduction}, we will describe this splitting ansatz in combination with the two schemes in Section \ref{sec:splitting_discretization}. Lastly in Section \ref{sec:num_exp}, numerical experiments will be presented, first as a verification of the AKW scheme, and after that the application to the full PyGyro model. Concluding with Section \ref{sec:conclusion}, we discuss the benefits of this approach. 
\end{abstract}

\begin{resume}%TODO: verify french
	En pratique, il existe plusieurs façons de résoudre les équations gyro-cinétiques. Ici, nous combinerons la méthode semi-lagrangienne (SL) et le schéma d'Arakawa (AKW) combiné à un intégrateur temporel. Les deux méthodes sont utilisées avec succès dans la pratique lorsqu'elles sont appliquées individuellement. Dans notre cas, nous les combinons en décomposant d'abord le problème en un système dynamique rapide (parallèle) et un système dynamique lent (perpendiculaire). L'approche SL sera utilisée pour résoudre le système rapide et le schéma AKW pour le sous-système lent. Jusqu'à présent, dans notre code de référence \cite{pygyro_code}, le modèle entier est résolu en utilisant uniquement la méthode SL. Notre objectif est de remplacer la discrétisation du sous-système lent par le schéma AKW, afin d'améliorer la préservation des quantités conservées du système physique en question. \\
	Après une introduction des équations gyro-cinétiques dans la Section \ref{sec:introduction}, nous décrirons cet ansatz de fractionnement en combinaison avec les deux schémas dans la Section \ref{sec:splitting_discretization}. Enfin, dans la section \ref{sec:num_exp}, des expériences numériques seront présentées, d'abord pour vérifier le schéma AKW, puis pour l'appliquer au modèle PyGyro complet. Dans la section \ref{sec:conclusion}, nous discutons des avantages de cette approche. 
\end{resume}

\maketitle
\newpage
\tableofcontents %TODO: can also be removed later

\input{sections/introduction.tex}

\input{sections/operator_splitting_and_discretization}

\input{sections/numerical_experiments}

\input{sections/conclusion}


\begin{acknowledgement}
	\textbf{Acknowledgements}\\
	The authors would like to thank Emmanuel Franck, Hélène Hivert, Guillaume Latu, Hélène Leman, Bertrand Maury, Michel Mehrenberger, and Laurent Navoret for organizing the CEMRACS conference 2022 and for the wonderful opportunity to come to Marseille and do research. We express special thanks to Michel Mehrenberger, Virginie Grandgirard, and Martin Campos Pinto for the daily supervision and general shaping of the project. Dominik Bell and Frederik Schnack thank Eric Sonnendrücker for the opportunity to participate in the CEMRACS and the fruitful discussions after the conference to give this project the final details. All authors are tremendously thankful for Emily Bourne and her help with the implementation and the work she did on the PyGyro code, which laid the ground-work for this project.
\end{acknowledgement}

\bibliographystyle{ieeetr}
\bibliography{literature}
\end{document}
